---
output:
  github_document
---

```{r, include = FALSE}
library(typewriter)

knitr::knit_hooks$set(error = function(x, options) { 
  paste(collapse = "\n", c(
    "```diff",
    paste0("- ", gsub("^## ", "", strsplit(x, "\n")[[1L]])), 
    "```"
  ))
})
```


# `typewriter`

An R package for runtime type-checking

## :scientist: Status: **Mad Science** :scientist:

This project is a thought exercise and syntactic experiment, trying to cobble
some typing syntax into R. It is not guaranteed to be robust, or to be well
vetted, but it does offer an introduction to type systems and concepts for
people familiar with R. 

For anything beyond educational use, use with caution. 

## :pencil2: Tutorial 

Since type systems can seem rather foreign in R, let's try to slowly build up
some motivating examples to show what value type systems add when developing R
code:

### Adding types to a function

Fundamentally, type systems are designed to impose some limits on what types of
data your function accepts. Practically all code assumes _something_ about your
inputs, and type systems provide a way of articulating those assumptions.

Specify types in the function header using `param = default :type` syntax.

```{r}
add <- type(function(x = 0 :numeric, y = 0 :numeric) {
  x + y
})

add(3, 14)
```

And if we try to give it something that isn't numeric, this is what we see:

```{r, error = TRUE}
add(3, "not a numeric")
```

### Using `%@%` to decorate a function with type checking

Since it can be a bit annoying to wrap functions like this, you can also use the
`%@%` operator (styled after python's `@` decorator syntax). 

> Functionally, this is a "reverse pipe"

```{r}
add <- type %@% function(x = 0 :numeric, y = 0 :numeric) {
  x + y
}
```

We'll use this style for the rest of the examples.

### Adding "traits"

R doesn't really have a concept of traits - behaviors that are defined over
groups of types. Testing for behavior, instead of defining behaviors by type is
typically referred to as an "interface". To do this in R, we can repurpose
functions that check behaviors of R objects. This includes many `is.*` or
`has_*` functions like `is.numeric` or `is.finite`.

```{r}
add_finite <- type %@% function(
  x = 0 :numeric(is.finite),
  y = 0 :numeric(is.finite)
) {
  x + y
}

add_finite(1, 2)
```

```{r, error = TRUE}
add_finite(1, NA)
```

Since we're repurposing functions that probably weren't intended to be used as
traits, we might need some extra flexibility to constrain their output. For
example, we might want to limit our adding function to only operate on vectors
with a specific length.

```{r}
add_numeric_vec <- type %@% function(
  x = 0 :numeric(length=10), 
  y = 0 :numeric(length=10)
) {
  x + y
}

add_numeric_vec(1:10, 11:20)
```

### Parameterized Types

Now, implementing an adding function for each length of vector would be pretty
tedious. What we really want, is to make sure that both input vectors have the
_same_ length, whatever that might be. For that we can use type parameters. 

```{r}
add_numeric_vec <- type(N) %@% function(
  x = 0 :numeric(length=N), 
  y = 0 :numeric(length=N)
) {
  x + y
}

add_numeric_vec(1:5, 2:6)
```

```{r, error = TRUE}
add_numeric_vec(1:5, 1:10)
```

### Generic Type

Finally, limiting our function so that it only applies to numeric types might
limit the use of other vectors that can be added, such as `complex` or
`difftime`. If all we need is for the type to implement the `+` operator, then
why should we need to re-write our function for each type of vector?

For this, we want a ***Generic*** type.

```{r}
add_vec <- type(T, N) %@% function(
  x = 0 :T(length=N), 
  y = 0 :T(length=N)
) {
  x + y
}

add_vec(1:5, 2:6)
```

```{r}
add_vec(complex(1:3, 2:4), complex(3:5, 4:6))
```

Cool! Our function is now _Generic_, only assuming that both inputs are the same
type and have the same length.

We can see how our function is different from `+` by looking at the behavior
when two different types are provided. `complex` and `numeric` vectors _can_ be
added together.

```{r}
complex(3, 1:3) + 3:5
```

But they fail when passed to our `add_vec` function because both inputs must
have the same type. Even though our function is _Generic_ over input types, it
still asserts constraints over the relationships between inputs.

```{r, error = TRUE}
add_vec(complex(3, 1:3), 3:5)
```

### Return Types (work in progress :construction:)

Just like function parameters, return types can be specified. The biggest
benefit to defining a return type is that a compiler or static type checker can
start to reason about the types of your variables. Until more type inferencing
is done based on return types, those benefits are not going to be realized, but
that isn't going to stop us from implementing it!

```{r, eval = FALSE}
add_vec <- type(T, N, return :T) %@% function(
  x = 0 :T(length=N),
  y = 0 :T(length=N)
) {
  x + y
}

add_vec(1:5, 2:6)
```

### Putting it all together

Many of the tools implemented here can be mixed and matched. For a full
break-down of all the syntax that is supported, take a look at `?type_match`,
the function that handles all the behind-the-scenes type checking. For now,
let's just take a look at some more involved examples.

#### `replace_similar`

Let's define a function for replacing elements in a list, but only when the new
values have the same type, so that we don't modify the overall structure of our
list.

``` {r}
replace_similar <- type(T, N) %@% function(
  df  = .:list[[at :T(length=N)]],
  at  = .:character(length=1),
  vec = .:T(length=N)
) {
  df[[at]] <- vec
  df
}
```

Let's give it a go 

```{r}
example_data <- list(a = 1:3, b = factor(c("a", "b")))
replace_similar(example_data, "b", factor(c("c", "d")))
```

This seems to work, but our factor levels _did_ change. This highlights one of
the challenges of type systems in a dynamic language. What exactly _is_ the type
of _factor_ anyways? Depending on how we want to draw this type-shaped box
around dynamically typed amorphous blob, we _could_ check that the attributes
are the same, but that may be overly specific for other data types. Since
there's no concensus, types in dynamic languages will inevitably be a bit loose.

Enough with the caveats, though - let's move on to more examples:

```{r}
mtmini <- mtcars[1:3, c("wt", "cyl")]
replace_similar(mtmini, "cyl", c(4, 5, 2))
```

Because a `data.frame` is just a `list` with style, our function works just as
well with a `data.frame` object. Even a `tibble` would work here. What
differentiates these objects isn't necessarily their type, but the interfaces
that they implement and how they implement them.

You may have also noticed that the `at` argument is actually used as part of the
type definition for the `df` param. This is some black magic wizardry that can
really only happen because we're doing run-time type checking. A type checking
algorithm would have to be pretty savvy to do these sorts of checks in static
analysis (before the code is executed).

## Other features

### Type Unions

Type definitions can also be unions using the `|` operator as a separator. 

```{r, eval = FALSE}
f <- type %@% function(x = .:numeric|complex) {
  mean(x)
}
```

### Indexed or Infixed Types

You can also apply type constraints to destructured elements. To type check
every element in a list you can check against `list[type]`, or for a specific
element, you can type check `list[["at" :type]]`. Likewise, you can specify that
the result of an infix function result using `rlang`-style lambda syntax such as
`list(.$a :numeric)`, or an alternative shorthand `list(a$numeric)`.

### Nested Structures

Type definitions may also define nested structures. Most likely, this type of
complex structure would be better checked by defining your own "interface"
function. Nevertheless, the checking is there if you want it.

```{r, eval = FALSE}
f <- type %@% function(
  t = .:list[["a" :list(length=3)[numeric], "b" :numeric]]
) {  
  sum(unlist(t$a), t$b)
}

f(list(a = list(1, 2, 3), b = c(4, 5, 6)))
```

## Further work

As it is now, this project is primarily just a playground for some syntax
exploration and language design musings. Depending on how it grows, there are a
few features that could use some attention (in order of likelyhood of actually
getting implemented):

- [ ] **Better type bounds**:  
  Instead of repeatedly applying the same type bounds anywhere where a T
  appears, it would be more convenient to constrain T in the type definition.

  ``` {r, eval = FALSE}
  replace_similar <- type(T(length=N)) %@% function(
    df  = .:list[[at :T]],
    at  = .:character(length=1),
    vec = .:T
  ) {
  }
  ```

- [ ] **Better error messages**:  
  The error messages are pretty terse and nondescript right now. It wouldn't be
  too heavy of a lift to output something more meaningful, especially for type
  parameters that failed to match where there's some relational qualities that
  are expected.

- [ ] **"Compiled" type checking**:  
  Currently, type checking does some gnarly non-standard evaluation, doing type
  checking by walking the type definition syntax tree with the input data each
  time the function is called. This introduces pretty considerable execution
  overhead. Instead, the checks could be "compiled" into assertions when the
  function is declared. This would make the code faster, and the function bodies
  mor intelligable. 

  If there's any interst in using this package beyond educational purposes, this
  would be the lowest hanging fruit to get it into reasonable shape for
  real-world use.

  In some cursory benchmarks comparing this run-time type analysis against
  hand-written assertions, compiled assertions ran ~3x faster. Naturally the
  gains are very dependent on the complexity of the type definitions.

- [ ] **Respecting Non-standard Evaluation**  
  R's non-standard evaluation throws a big wrench into this whole thing. There's
  currently no good way to do type checking on arguments that aren't intended to
  be evaluated. For example, there's no good way to define a type such as:

  ```{r, eval = FALSE}
  code_echo <- type %@% function(expr = .:expression[1: "{"]) {
  }
  ```

  Maybe that's a good thing? Just leave it untyped if you plan to do something
  wild with it. Otherwise, I can see this being quite the rabbit hole to fall
  into.

- [ ] **Return types**:  
  Without something that actually uses these types, return types aren't
  particulary valuable. Nevertheless, what type of type system doesn't have
  return types!

  Unless a tools is interested in making use of return types for some static
  type checking, this is quite low priority.
