---
title: "Writing Typed Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{writing_typed_functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(typing)
```

# Introduction

R is a dynamically typed language, and it would be an incredible effort to try
to introduce truly static typing behaviors into R and robustly handle the
incredible flexibility that non-standard evaluation offers. 

Instead, typing can be used to perform type assertions at runtime. Often this
comes in the form of defensive programming and assertions. This package simply
aims to make those practices easier to embed and easier to integrate into a
package development workflow. 

To do this, we pass a "typed" function definition to the `type` function.

```{r introduction}
type(function(a = 1:10 :numeric) {
  sum(a)
})
```

As we can see, the type function does some clean-up of our function. Even though
the original function was syntactically _valid_, it would most likely throw an
error. The `type` function takes type definitions and embeds them as a type
checking function call within the function body. There's still some non-standard
stuff tucked into the `type_constrain` and `type_check` function calls, but at
least our function header looks more familiar.

# Type Checking

Type checking is handled by the `type_check` function. This function looks at
the surrounding environment and checks available values against type
definitions.

```{r error}
with(list(a = 1L), type_check(c(), a = character))
```

> For now we're going to ignore the first argument and focus on the remaining
> named arguments. For the following examples, we'll ignore the outer `with()`
> call.

## Type Definitions

There are a few different ways that types can be defined as:

1. `character`  
   A `character` type is considered to be satisfied if the object inherits a
   matching class, or if `mode(x)` evaluates to the `character` value (e.g. `1L`
   will satisfy the type definition `"numeric"`).

   **Examples:** 
   ```
   type_check(c(), a = "character")
   ```

1. `name`  
   If a predicate function (returning a singular logical value) exists in the
   function definition environment, then an argument is considered to satisfy
   the type definition if the predicate returns `TRUE`. If no such predicate
   function exists, the name is interpretted as though it is a `character` type
   definition.

   **Examples:** 
   ```
   type_check(c(), a = numeric)     # character
   type_check(c(), a = is.numeric)  # predicate
   ```

1. `function`
   Similar to the predicates provided as a `name`, predicates can be provided as
   in in-line function.

   **Examples:** 
   ```
   type_check(c(), a = function(i) length(i) == 3L)
   ```

1. `call`
   Although represented like a call, this syntax is meant to be used to define
   type traits (or rather, interfaces). You may want to accept arguments that
   are numeric vectors of a particular length. In this way, you can provide
   interfaces and the expected output for your type bounds. You could define a
   custom predicate to constrain your type, or use the `call` syntax to apply an
   in-line constraint. This behavior is easier shown:

   **Examples:**
   ```
   type_check(c(), a = numeric(length=3L))
   type_check(c(), a = list(names=c("x", "y")))
   ```

1. Unions
   This mode of definition is a special syntactic case, using the `|` operator
   as a shorthand for a type union, allowing for an argument to be any of the
   provided types. Each type in the union is evaluated separately and can use
   any of the syntaxes above.

   **Examples:**
   ```
   type_check(c(), a = character|factor)
   ```

## Type Parameters

Many type systems implement a form of generalization which uses type parameters
to constrain valid arguments.

For example, if we want to make a function that produces a `data.frame`, but
want to avoid implicit recycling of column values, then we may want to insist
that all arguments have the same length. This is a perfect use case for a type
parameter. For the sake of example, we'll also insist that all columns are less
than 5 rows long.

```{r}
with(list(x = 1:3, y = 2:4), { 
  type_constrain(type_check(c("T", "N"), x = T(length=N), y = T(length=N)), N < 5)
})
```

# Syntactic Sugar

Although you are free to check types using the `type_check` and `type_constrain`
functions yourself, the `type` and `where` functions are provided to process
functions that use a type definition syntactic sugar.

```{r}
type(function(a = 1:3 :numeric, b = letters[1:3] :character) {
  print(paste(a, b))
})
```

The contents after the last `:` are used as the type signature, and embedded
within the function to perform type checks.

> **Note**
>
> If you'd like to provide a type for an argument that has no default value, you
> must provide a placeholder `.` in order to use this type syntax.
>
> ```
> function(a = . :character) {}
> ```

## Type Parameters

Likewise, the `where` function can be used to set type parameters.

